### 2020-02-07 ncov_lancet_ID.tex
crap

\usepackage{natbib}
\setcitestyle{numbers} 
\setcitestyle{square}
\bibliographystyle{unsrt}
\date{\today}

@@

\begin{flushleft}{
	\Large
	\textbf\newline{
		Reconciling early-outbreak estimates of the basic reproductive number and its uncertainty: framework and applications to the novel coronavirus (2019-nCoV) outbreak
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,*}
Benjamin M.\ Bolker\textsuperscript{2,3,4}
David Champredon\textsuperscript{5}
David J.\,D.\ Earn\textsuperscript{3,4}
Michael Li\textsuperscript{2}
Joshua S.\ Weitz\textsuperscript{6, 7}
Bryan T.\ Grenfell\textsuperscript{1,8,9}
Jonathan Dushoff\textsuperscript{2,3,4,*}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} Department of Biology, McMaster University, Hamilton, ON, Canada
\\
\textbf{3} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
\\
\textbf{4} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
\\
\textbf{5} Department of Pathology and Laboratory Medicine, University of Western Ontario, London, Ontario, Canada
\\
\textbf{6} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{7} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{8} Division of International Epidemiology and Population Studies, Fogarty International Center, National Institutes of Health, Bethesda, MD, USA
\\
\textbf{9} Woodrow Wilson School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\bigskip

@@

We decompose \Ro into three key quantities: the exponential growth rate $r$, the mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$.
We use a Bayesian multilevel model to construct pooled estimates and measure uncertainties associated with these quantities.
We assess early estimates of \Ro for the 2019-nCoV outbreak using the pooled estimates.

@@

7 analyses of the 2019-nCoV outbreak were published online as pre-prints between January 23--26, 2020.
Their modeling approaches varied widely and estimates of uncertainties in \Ro ranged from 1.5 to 7.1.

@@

Since December 2019 \citep{pneumonia}, a novel coronavirus (2019-nCoV) has been spreading in China and other parts of the world.
Although the virus is believed to have originated from animal reservoirs, the ability of 2019-nCoV to directly transmit between humans has posed a greater threat for its spread \citep{chan2020familial}.
As of February 3, 2020, the World Health Organization (WHO) has confirmed 17391 cases, including 153 confirmed cases in 23 different countries, outside China \citep{who02report}.
% Also on January 30, 2020, WHO declared the outbreak a Public Health Emergency of International Concern \citep{whoemer}.

As the disease continues to spread, many researchers have published their preliminary analyses of the outbreak as pre-prints (e.g., \cite{bedfordncov, imaincov, liuncov, majumderncov, readncov, zhaoncov}), focusing in particular on estimates of the basic reproductive number \Ro (i.e., the average number of secondary cases generated by a primary case in a fully susceptible population \citep{anderson1991infectious}).
Estimates of the basic reproductive number are of interest during an outbreak because they provide information about the level of intervention required to interrupt transmission, and about the final size of the outbreak \citep{anderson1991infectious}.
We commend these researchers for their timely contribution and those who made the data publicly available.
However, it can be difficult to compare a disparate set of estimates of \Ro from different research groups when the estimation methods and their underlying assumptions vary widely.

Here, we show that a wide range of approaches to estimating \Ro can be understood and compared in terms of estimates of three quantities: the exponential growth rate $r$, the mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$.
The generation interval, defined as the interval between the time when an individual becomes infected and the time when that individual infects another individual \citep{svensson2007note}, plays a key role in shaping the relationship between $r$ and \Ro \citep{wallinga2007generation};
therefore, estimates of \Ro from different models directly depend on their implicit assumptions about the generation-interval distribution and the exponential growth rate.
Early in an epidemic, information is scarce and there is inevitably a great deal of uncertainty surrounding disease transmission.
We suggest that disease modelers should make sure their assumptions about these three quantities are clear and reasonable, and that estimates of uncertainty in \Ro should propagate error from all three sources \citep{elderd2006uncertainty}.

We compare seven disparate models published online between January 23--26, 2020 that estimated \Ro for the 2019-nCoV outbreak \citep{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}. 
We use our framework to construct pooled estimates for the three key quantities: $r$, $\bar G$, and $\kappa$;
the pooled estimates reflect the uncertainties present in modeling approaches.
We use these pooled estimates to illustrate the importance of propagating different sources of error, particularly uncertainty in both the growth rate and the generation interval.
We also use our framework to tease apart which assumptions of these different models led to their different estimates and confidence intervals. 
Despite the availability of more recent and/or updated estimates of \Ro, we restrict ourselves to the estimates above in order to focus on the resolution of uncertainty in the earliest stages of an epidemic.

@@

We gathered information on estimates of \Ro and their assumptions about the underlying generation-interval distributions from 7 articles that were published online between January 23--26, 2020 (Table 1).
Five studies \citep{liuncov, majumderncov, readncov, riouncov, zhaoncov} were uploaded to pre-print servers (bioRxiv, medRxiv, and SSRN); one report was posted on the web site of Imperial College London \citep{imaincov}; and one report was posted on nextstrain.org \citep{bedfordncov}.
They vary widely in their modeling approaches: a branching process model \citep{bedfordncov, imaincov, riouncov}, a deterministic Susceptible-Exposed-Infected-Recovered (SEIR) model \citep{readncov}, an exponential growth model \citep{zhaoncov}, a Poisson offspring distribution model \citep{liuncov}, and the Incidence Decay and Exponential Adjustment (IDEA) model \citep{majumderncov}.
Four studies estimated \Ro by directly fitting their models to incidence data \citep{readncov, zhaoncov, liuncov, majumderncov}.
The remaining three studies estimated \Ro by comparing the predicted number of cases from their models with the estimated number of total cases by January 18 (between 1,000 and 9,7000 \citep{imaincov0})
Some of these studies have now been published in peer-reviewed journals \citep{riou2020pattern, zhao2020preliminary} or have been updated with better uncertainty quantification \citep{readncov2}.

@@

Early in an outbreak, \Ro is difficult to estimate directly;
instead, \Ro is often inferred from the exponential growth rate $r$, which can be estimated reliably from incidence data \citep{ma2014estimating}.
Given an estimate of the exponential growth rate $r$ and an \emph{intrinsic} generation-interval distribution $g(\tau)$ \citep{champredon2015intrinsic}, the basic reproductive
number can be estimated via the Euler-Lotka equation \citep{wallinga2007generation}:
\begin{linenomath*}
\begin{equation}
1/\Ro = \int \exp(-r\tau) g(\tau) \, d\tau.
\label{eq:euler}
\end{equation}
\end{linenomath*}
In other words, estimates of \Ro must
depend on the assumptions about the
exponential growth rate $r$ and the shape of the generation-interval distribution $g(\tau)$.

Here, we use the gamma approximation framework \citep{park2019practical} to (i) characterize the amount of uncertainty present in the exponential growth rates and the shape of the generation-interval distribution and (ii) assess the degree to which these uncertainties affect the estimate of \Ro.
Assuming that generation intervals follow a gamma distribution
with the mean $\bar G$ and the squared coefficient of variation $\kappa$, 
we have
\begin{linenomath*}
\begin{equation}
\Ro = \left(1 + \kappa r \bar{G}\right)^{1/\kappa}.
\label{eq:gamma}
\end{equation}
\end{linenomath*}
This equation demonstrates that a generation-interval distribution
that has a larger mean (higher $\bar{G}$) or is less variable (lower $\kappa$)
will give a higher estimate of \Ro for the same value of $r$.

@@

\newcommand{\gammdist}{\mathrm{Gamma}}
\begin{table}[t]
\begin{center}
\scriptsize
\begin{tabular}{l|p{4.5cm}|p{2.5cm}|p{2.7cm}}
 & Basic reproductive\newline number \Ro & Mean generation\newline interval $\bar G$ (days) & Generation-interval\newline dispersion $\kappa$ \\
\hline
Study 1 & $\mathrm{Uniform}(1.5, 3.5)$ & 10 & 1 \\
\hline
Study 2 & $\gammdist(\mathrm{mean}=2.6, \alpha=28)$ & 8.4 & 0.5 \\
\hline
Study 3 & $\gammdist(\mathrm{mean}=2.92, \alpha=67)$ & 8.4 & 0.2 \\
\hline
Study 4 & $\gammdist(\mathrm{mean}=3.8, \alpha=1400)$ & 7.6 & 0.5 \\
\hline
Study 5 & $\gammdist(\mathrm{mean}=2.2, \alpha=12)$ & $\mathrm{Uniform}(7, 14)$ & 0.5\\
\hline
Study 6 & $\gammdist(\mathrm{mean}=5.47, \alpha=54)$ & $\mathrm{Uniform}(7.6, 8.4)^\ast$ & 0.2\\
\hline
Study 7 & $\exp(r \bar G)^\dagger$ & $\mathrm{Uniform}(6, 10)$ & 0\\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Probability distributions for \Ro, $\bar G$, and $\kappa$.}
We use these probability distributions to obtain a probability distribution for the exponential growth rate $r$.
The gamma distribution is parameterized by its mean and shape $\alpha$.
Constant values are fixed according to Table 1.
$^\ast$We do not account for this uncertainty during our recalculation of the exponential growth rate $r$ because the reported estimate of $\mathcal R_0$ and its uncertainty assumes $\bar G = 8$; the original article reports three $\mathcal R_0$ (and 95\% CIs) estimates using three different values of $\bar G$: 7.6 (MERS-like), 8 (average), and 8.4 (SARS-like).
We still account for this uncertainty in our pooled estimates ($\mu_G$).
$^\dagger$Study 6 uses the IDEA model \citep{fisman2013idea}, through which the authors effectively fit an exponential curve to the cumulative number of confirmed cases without propagating any statistical uncertainty.
Instead of modeling \Ro with a probability distribution and recalculating $r$, we use $r=0.114\,\mathrm{days}^{-1}$, which explains all uncertainty in the reported \Ro, when combined with the considered range of $\bar G$.
}
\end{table}

We construct pooled estimates for each parameter ($r$, $\bar G$, and $\kappa$) using a Bayesian multilevel modeling approach, which assumes that the parameters across different studies come from the same gamma distribution:
\begin{linenomath*}
\begin{equation}
\begin{aligned}
r_i &\sim \gammdist(\mathrm{mean}=\mu_r, \mathrm{shape}=\mu_r^2/\sigma_r^2),\\
\bar{G}_i &\sim \gammdist(\mathrm{mean}=\mu_G, \mathrm{shape}=\mu_G^2/\sigma_G^2),\\
\kappa_i &\sim \gammdist(\mathrm{mean}=\mu_\kappa, \mathrm{shape}=\mu_\kappa^2/\sigma_\kappa^2),\\
\end{aligned}
\end{equation}
\end{linenomath*}
where $\sigma_r$, $\sigma_G$, and $\sigma_\kappa$ represent between-study standard deviations in $r$, $\bar G$, and $\kappa$, respectively.
We account for uncertainties associated with $r_i$, $\bar G_i$ and $\kappa_i$ (and their correlations), by drawing a random set from the family of parameter sets for each study at each Metropolis-Hastings step.
Since the gamma distribution does not allow zeros, we use $\kappa=0.02$ instead for Study 7.
We note that this approach does not account for non-independence between the parameter estimates made by different modelers.
As we add more models, the pooled estimates can become sharper even when the models no longer add more information.
Thus, the pooled estimator should be interpreted with care.

We use weakly informative priors on hyperparameters representing pooled means and non-informative priors on hyperparameters representing between-study standard deviations:
\begin{linenomath*}
\begin{equation}
\begin{aligned}
\mu_r &\sim \gammdist(\mathrm{mean}=1\,\mathrm{week}^{-1},\,\mathrm{shape}=2)\\
\mu_G &\sim \gammdist(\mathrm{mean}=1\,\mathrm{week},\,\mathrm{shape}=2)\\
\mu_\kappa &\sim \gammdist(\mathrm{mean}=0.5,\,\mathrm{shape}=2)\\
(\sigma_r, \sigma_G, \sigma_\kappa) &\sim \textrm{half-normal}(0, 10).
\end{aligned}
\end{equation}
\end{linenomath*}
We followed recommendations outlined in Gelman \textit{et al.} \cite{gelman2006prior}, parameterizing the top-level gamma distributions in terms of their means and standard deviations and imposing weakly informative prior distributions on between-study standard deviations, i.e., $\textrm{half-normal}(0, 10)$.
We had initially used gamma priors with small shape parameters ($< 1$) on between-study shape parameters ($=\mu^2/\sigma^2$) but found this put too much prior probability on large between-study variances. This phenomenon is a known problem \citep{gelman2006prior}.
Alternative choices of prior for the between-study shape parameters are also suboptimal: imposing strong priors (e.g. half-Student-$t(\mu=0,\sigma=1,\nu=4)$  assumes \textit{a priori} that between-study variance is large,  while weak priors (e.g. half-Cauchy(0,5)) can lead to poor mixing.

We run 4 independent Markov Chain Monte Carlo chains each consisting of 500,000 burnin steps and 500,000 sampling steps.
Posterior samples are thinned every 1000 steps.
Convergence is assessed by ensuring that the Gelman-Rubin statistic is below 1.01 for all hyperparameters \citep{gelman1992inference};
trace plots and marginal posterior distribution plots are presented in Appendix.
95\% confidence intervals are calculated by taking 2.5\% and 97.5\% quantiles from the marginal posterior distribution for each parameter.
\texttt{R} code is available in GitHub (\url{https://github.com/parksw3/nCoV_framework}).

@@

\fref{assumption} compares the reported values of the exponential growth rate $r$, mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$ from different studies with the pooled estimates that we calculate from our multilevel model.
We find that there is a large uncertainty associated with the underlying parameters;
many models rely on stronger assumptions that ignore these uncertainties.
Surprisingly, no studies take into account how the variation in generation intervals affects their estimates of \Ro:
all studies assumed fixed values for $\kappa$, ranging from 0 to 1.
% Assuming fixed parameter values can lead to overly strong conclusions \citep{elderd2006uncertainty}.

@@

\fref{eff} shows how propagating uncertainty in different combinations would affect estimates and CIs for \Ro. For illustrative purposes, we use our pooled estimates, which may represent a reasonable proxy for the state of knowledge as of January 23--26 (\fref{eff}A).
Comparing the models that include only some sources of uncertainty to the \textbf{all} model, we see that propagating error from the growth rate (which all but one of the studies reviewed did) is absolutely crucial: the middle bar (\textbf{GI mean}), which lacks growth-rate uncertainty, is relatively narrow.
In this case, propagating error from the mean generation interval has negligible effect compared to propagating the uncertainty in $r$.
Uncertainty in the generation-interval dispersion also has important effects as it determines the functional form of the relationship between $r$ and \Ro (compare \textbf{growth rate + GI mean} with \textbf{all}).
For example, reducing the dispersion parameter $\kappa$ from 1 (assuming exponentially distributed generation intervals) to 0 (assuming fixed generation intervals) changes the $r$--\Ro\ relationship from linear to exponential, therefore increasing the sensitivity of \Ro estimates to $r$ and $\bar G$.

@@

We also compare the estimates of \Ro across different studies by 
replacing their values of $r$, $\bar G$, and $\kappa$ with our pooled estimates ($\mu_r$, $\mu_G$, and $\mu_\kappa$, respectively) one at a time and recalculating the basic reproductive number \Ro (\fref{R0}).
This procedure allows us to assess the sensitivity of the estimates of \Ro across appropriate ranges of uncertainties.
We find that incorporating uncertainties one at a time increases the width of the confidence intervals in all but 7 cases.
We estimate narrower confidence intervals for Study 3, Study 6, and Study 7 when we account for proper uncertainties in the generation-interval dispersion because they assume a narrow generation-interval distribution (compare \textbf{base} with \textbf{GI variation} in \fref{R0});
when higher values of $\kappa$ are used, their estimates of \Ro become less sensitive to the values of $r$ and $\bar G$, giving narrower confidence intervals.
We estimate narrower confidence intervals for Study 5 and Study 7 when we account for proper uncertainties in the mean generation interval (compare \textbf{base} with \textbf{GI mean} in \fref{R0}) because the range of uncertainty in the mean generation interval $\bar G$ they consider is much wider than the pooled range (\fref{assumption}).
Substituting the reported $r$ or $\bar G$ from Study 1 with our pooled estimates give narrower confidence intervals for similar reasons.

@@

Estimating the basic reproductive number \Ro is crucial for predicting the course of an outbreak and planning intervention strategies.
Here, we use a gamma approximation \citep{park2019practical} to decompose \Ro estimates into three key quantities ($r$, $\bar G$, and $\kappa$) and apply a multilevel Bayesian framework to compare estimates of \Ro for the novel coronavirus outbreak.
Our results demonstrate the importance of accounting for uncertainties associated with the underlying generation-interval distributions, including uncertainties in the amount of dispersion in the generation intervals:
our analysis of individual studies shows that assuming too narrow a generation-interval distribution can make the estimate of \Ro overly sensitive to the estimates of the exponential growth rate $r$.

Of the seven studies that we reviewed, two of them directly fit their models to cumulative number of confirmed cases.
This approach can be appealing because of its simplicity and apparent robustness, but fitting a model to cumulative incidence instead of raw incidence can both bias parameters and give overly narrow confidence intervals, if the resulting non-independent error structures are not taken into account \citep{ma2014estimating, king2015avoidable}.
Naive fits to cumulative incidence data should therefore be avoided.

>>>>

\usepackage{natbib}
\usepackage[superscript,biblabel]{cite}
\setcitestyle{numbers} 
\setcitestyle{square}
\bibliographystyle{unsrt}
\date{\today}

@@

\begin{flushleft}{
	\Large
	\textbf\newline{
		Reconciling early-outbreak estimates of the basic reproductive number and its uncertainty: framework and applications to the novel coronavirus (2019-nCoV) outbreak
	}
}
\newline
\\
Sang Woo Park BSc\textsuperscript{1,*}
Benjamin M.\ Bolker PhD\textsuperscript{2,3,4}
David Champredon PhD\textsuperscript{5}
David J.\,D.\ Earn PhD\textsuperscript{3,4}
Michael Li PhD\textsuperscript{2}
Joshua S.\ Weitz PhD\textsuperscript{6, 7}
Bryan T.\ Grenfell PhD\textsuperscript{1,8,9}
Jonathan Dushoff PhD\textsuperscript{2,3,4,*}
\\
\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, NJ, USA
\\
\textbf{2} Department of Biology, McMaster University, Hamilton, ON, Canada
\\
\textbf{3} Department of Mathematics and Statistics, McMaster University, Hamilton, ON, Canada
\\
\textbf{4} M.\,G.\,DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, ON, Canada
\\
\textbf{5} Department of Pathology and Laboratory Medicine, University of Western Ontario, London, Ontario, Canada
\\
\textbf{6} School of Biological Sciences, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{7} School of Physics, Georgia Institute of Technology, Atlanta, GA, USA
\\
\textbf{8} Division of International Epidemiology and Population Studies, Fogarty International Center, National Institutes of Health, Bethesda, MD, USA
\\
\textbf{9} Woodrow Wilson School of Public and International Affairs, Princeton University, Princeton, NJ, USA
\\
\bigskip

@@

We reviewed 7 studies that were published online between January 23--26, 2020.
We decompose their \Ro estimates into three key quantities: the exponential growth rate $r$, the mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$.
We use a Bayesian multilevel model to construct pooled estimates and measure uncertainties associated with these quantities.
We assess early estimates of \Ro for the 2019-nCoV outbreak using the pooled estimates.

@@

% 7 analyses of the 2019-nCoV outbreak were published online as pre-prints between January 23--26, 2020.
% Their modeling approaches varied widely and estimates of uncertainties in \Ro ranged from 1.5 to 7.1.

@@

Since December 2019 \cite{pneumonia}, a novel coronavirus (2019-nCoV) has been spreading in China and other parts of the world.
Although the virus is believed to have originated from animal reservoirs, the ability of 2019-nCoV to directly transmit between humans has posed a greater threat for its spread \cite{chan2020familial}.
As of February 3, 2020, the World Health Organization (WHO) has confirmed 17391 cases, including 153 confirmed cases in 23 different countries, outside China \cite{who02report}.
% Also on January 30, 2020, WHO declared the outbreak a Public Health Emergency of International Concern \cite{whoemer}.

As the disease continues to spread, many researchers have published their preliminary analyses of the outbreak as pre-prints (e.g., \cite{bedfordncov, imaincov, liuncov, majumderncov, readncov, zhaoncov}), focusing in particular on estimates of the basic reproductive number \Ro (i.e., the average number of secondary cases generated by a primary case in a fully susceptible population \cite{anderson1991infectious}).
Estimates of the basic reproductive number are of interest during an outbreak because they provide information about the level of intervention required to interrupt transmission, and about the final size of the outbreak \cite{anderson1991infectious}.
We commend these researchers for their timely contribution and those who made the data publicly available.
However, it can be difficult to compare a disparate set of estimates of \Ro from different research groups when the estimation methods and their underlying assumptions vary widely.

Here, we show that a wide range of approaches to estimating \Ro can be understood and compared in terms of estimates of three quantities: the exponential growth rate $r$, the mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$.
The generation interval, defined as the interval between the time when an individual becomes infected and the time when that individual infects another individual \cite{svensson2007note}, plays a key role in shaping the relationship between $r$ and \Ro \cite{wallinga2007generation};
therefore, estimates of \Ro from different models directly depend on their implicit assumptions about the generation-interval distribution and the exponential growth rate.
Early in an epidemic, information is scarce and there is inevitably a great deal of uncertainty surrounding disease transmission.
We suggest that disease modelers should make sure their assumptions about these three quantities are clear and reasonable, and that estimates of uncertainty in \Ro should propagate error from all three sources \cite{elderd2006uncertainty}.

We compare seven disparate models published online between January 23--26, 2020 that estimated \Ro for the 2019-nCoV outbreak \cite{bedfordncov, imaincov, liuncov, majumderncov, readncov, riouncov, zhaoncov}. 
We use our framework to construct pooled estimates for the three key quantities: $r$, $\bar G$, and $\kappa$;
the pooled estimates reflect the uncertainties present in modeling approaches.
We use these pooled estimates to illustrate the importance of propagating different sources of error, particularly uncertainty in both the growth rate and the generation interval.
We also use our framework to tease apart which assumptions of these different models led to their different estimates and confidence intervals. 
Despite the availability of more recent and/or updated estimates of \Ro, we restrict ourselves to the estimates above in order to focus on the resolution of uncertainty in the earliest stages of an epidemic.

@@

We gathered information on estimates of \Ro and their assumptions about the underlying generation-interval distributions from 7 articles that were published online between January 23--26, 2020 (Table 1).
Five studies \cite{liuncov, majumderncov, readncov, riouncov, zhaoncov} were uploaded to pre-print servers (bioRxiv, medRxiv, and SSRN); one report was posted on the web site of Imperial College London \cite{imaincov}; and one report was posted on nextstrain.org \cite{bedfordncov}.
Their modeling approaches vary widely: a branching process model \cite{bedfordncov, imaincov, riouncov}, a deterministic Susceptible-Exposed-Infected-Recovered (SEIR) model \cite{readncov}, an exponential growth model \cite{zhaoncov}, a Poisson offspring distribution model \cite{liuncov}, and the Incidence Decay and Exponential Adjustment (IDEA) model \cite{majumderncov}.
Four studies estimated \Ro by directly fitting their models to incidence data \cite{readncov, zhaoncov, liuncov, majumderncov}.
The remaining three studies estimated \Ro by comparing the predicted number of cases from their models with the estimated number of total cases by January 18 (between 1,000 and 9,7000 \cite{imaincov0})
Some of these studies have now been published in peer-reviewed journals \cite{riou2020pattern, zhao2020preliminary} or have been updated with better uncertainty quantification \cite{readncov2}.

@@

Early in an outbreak, \Ro is difficult to estimate directly;
instead, \Ro is often inferred from the exponential growth rate $r$, which can be estimated reliably from incidence data \cite{ma2014estimating}.
Given an estimate of the exponential growth rate $r$ and an \emph{intrinsic} generation-interval distribution $g(\tau)$ \cite{champredon2015intrinsic}, the basic reproductive
number can be estimated via the Euler-Lotka equation \cite{wallinga2007generation}:
\begin{linenomath*}
\begin{equation}
1/\Ro = \int \exp(-r\tau) g(\tau) \, d\tau.
\label{eq:euler}
\end{equation}
\end{linenomath*}
In other words, estimates of \Ro must
depend on the assumptions about the
exponential growth rate $r$ and the shape of the generation-interval distribution $g(\tau)$.

Here, we use the gamma approximation framework \cite{park2019practical} to (i) characterize the amount of uncertainty present in the exponential growth rates and the shape of the generation-interval distribution and (ii) assess the degree to which these uncertainties affect the estimate of \Ro.
Assuming that generation intervals follow a gamma distribution
with the mean $\bar G$ and the squared coefficient of variation $\kappa$, 
we have
\begin{linenomath*}
\begin{equation}
\Ro = \left(1 + \kappa r \bar{G}\right)^{1/\kappa}.
\label{eq:gamma}
\end{equation}
\end{linenomath*}
This equation demonstrates that a generation-interval distribution
that has a larger mean (higher $\bar{G}$) or is less variable (lower $\kappa$)
will give a higher estimate of \Ro for the same value of $r$.

@@

\newcommand{\gammdist}{\mathrm{Gamma}}
\begin{table}[t]
\begin{center}
\scriptsize
\begin{tabular}{l|p{4.5cm}|p{2.5cm}|p{2.7cm}}
 & Basic reproductive\newline number \Ro & Mean generation\newline interval $\bar G$ (days) & Generation-interval\newline dispersion $\kappa$ \\
\hline
Study 1 & $\mathrm{Uniform}(1.5, 3.5)$ & 10 & 1 \\
\hline
Study 2 & $\gammdist(\mathrm{mean}=2.6, \alpha=28)$ & 8.4 & 0.5 \\
\hline
Study 3 & $\gammdist(\mathrm{mean}=2.92, \alpha=67)$ & 8.4 & 0.2 \\
\hline
Study 4 & $\gammdist(\mathrm{mean}=3.8, \alpha=1400)$ & 7.6 & 0.5 \\
\hline
Study 5 & $\gammdist(\mathrm{mean}=2.2, \alpha=12)$ & $\mathrm{Uniform}(7, 14)$ & 0.5\\
\hline
Study 6 & $\gammdist(\mathrm{mean}=5.47, \alpha=54)$ & $\mathrm{Uniform}(7.6, 8.4)^\ast$ & 0.2\\
\hline
Study 7 & $\exp(r \bar G)^\dagger$ & $\mathrm{Uniform}(6, 10)$ & 0\\
\hline
\end{tabular}
\end{center}
\caption{
\textbf{Probability distributions for \Ro, $\bar G$, and $\kappa$.}
We use these probability distributions to obtain a probability distribution for the exponential growth rate $r$.
The gamma distribution is parameterized by its mean and shape $\alpha$.
Constant values are fixed according to Table 1.
$^\ast$We do not account for this uncertainty during our recalculation of the exponential growth rate $r$ because the reported estimate of $\mathcal R_0$ and its uncertainty assumes $\bar G = 8$; the original article reports three $\mathcal R_0$ (and 95\% CIs) estimates using three different values of $\bar G$: 7.6 (MERS-like), 8 (average), and 8.4 (SARS-like).
We still account for this uncertainty in our pooled estimates ($\mu_G$).
$^\dagger$Study 6 uses the IDEA model \cite{fisman2013idea}, through which the authors effectively fit an exponential curve to the cumulative number of confirmed cases without propagating any statistical uncertainty.
Instead of modeling \Ro with a probability distribution and recalculating $r$, we use $r=0.114\,\mathrm{days}^{-1}$, which explains all uncertainty in the reported \Ro, when combined with the considered range of $\bar G$.
}
\end{table}

We construct pooled estimates for each parameter ($r$, $\bar G$, and $\kappa$) using a Bayesian multilevel modeling approach, which assumes that the parameters across different studies come from the same gamma distribution:
\begin{linenomath*}
\begin{equation}
\begin{aligned}
r_i &\sim \gammdist(\mathrm{mean}=\mu_r, \mathrm{shape}=\mu_r^2/\sigma_r^2),\\
\bar{G}_i &\sim \gammdist(\mathrm{mean}=\mu_G, \mathrm{shape}=\mu_G^2/\sigma_G^2),\\
\kappa_i &\sim \gammdist(\mathrm{mean}=\mu_\kappa, \mathrm{shape}=\mu_\kappa^2/\sigma_\kappa^2),\\
\end{aligned}
\end{equation}
\end{linenomath*}
where $\mu_r$, $\mu_G$, $\mu_\kappa$ represent the pooled estimates, and $\sigma_r$, $\sigma_G$, and $\sigma_\kappa$ represent between-study standard deviations.
We account for uncertainties associated with $r_i$, $\bar G_i$ and $\kappa_i$ (and their correlations), by drawing a random set from the family of parameter sets for each study at each Metropolis-Hastings step.
Since the gamma distribution does not allow zeros, we use $\kappa=0.02$ instead for Study 7.
We note that this approach does not account for non-independence between the parameter estimates made by different modelers.
As we add more models, the pooled estimates can become sharper even when the models no longer add more information.
Thus, the pooled estimator should be interpreted with care.

We use weakly informative priors on hyperparameters:
\begin{linenomath*}
\begin{equation}
\begin{aligned}
\mu_r &\sim \gammdist(\mathrm{mean}=1\,\mathrm{week}^{-1},\,\mathrm{shape}=2)\\
\mu_G &\sim \gammdist(\mathrm{mean}=1\,\mathrm{week},\,\mathrm{shape}=2)\\
\mu_\kappa &\sim \gammdist(\mathrm{mean}=0.5,\,\mathrm{shape}=2)\\
(\sigma_r, \sigma_G, \sigma_\kappa) &\sim \textrm{half-normal}(0, 10).
\end{aligned}
\end{equation}
\end{linenomath*}
We followed recommendations outlined in Gelman \textit{et al.} \cite{gelman2006prior}, parameterizing the top-level gamma distributions in terms of their means and standard deviations and imposing weakly informative prior distributions on between-study standard deviations, i.e., $\textrm{half-normal}(0, 10)$.
We had initially used gamma priors with small shape parameters ($< 1$) on between-study shape parameters ($=\mu^2/\sigma^2$) but found this put too much prior probability on large between-study variances. This phenomenon is a known problem \cite{gelman2006prior}.
Alternative choices of prior for the between-study shape parameters are also suboptimal: imposing strong priors (e.g. half-Student-$t(\mu=0,\sigma=1,\nu=4)$  assumes \textit{a priori} that between-study variance is large,  while weak priors (e.g. half-Cauchy(0,5)) can lead to poor mixing.

We run 4 independent Markov Chain Monte Carlo chains each consisting of 500,000 burnin steps and 500,000 sampling steps.
Posterior samples are thinned every 1000 steps.
Convergence is assessed by ensuring that the Gelman-Rubin statistic is below 1.01 for all hyperparameters \cite{gelman1992inference};
trace plots and marginal posterior distribution plots are presented in Appendix.
95\% confidence intervals are calculated by taking 2.5\% and 97.5\% quantiles from the marginal posterior distribution for each parameter.
\texttt{R} code is available in GitHub (\url{https://github.com/parksw3/nCoV_framework}).

@@

\fref{assumption} compares the reported values of the exponential growth rate $r$, mean generation interval $\bar G$, and the generation-interval dispersion $\kappa$ from different studies with the pooled estimates that we calculate from our multilevel model.
We find that there is a large uncertainty associated with the underlying parameters;
many models rely on stronger assumptions that ignore these uncertainties.
Surprisingly, no studies take into account how the variation in generation intervals affects their estimates of \Ro:
all studies assumed fixed values for $\kappa$, ranging from 0 to 1.
% Assuming fixed parameter values can lead to overly strong conclusions \cite{elderd2006uncertainty}.

@@

\fref{eff} shows how propagating uncertainty in different combinations would affect estimates and CIs for \Ro. For illustrative purposes, we use our pooled estimates, which may represent a reasonable proxy for the state of knowledge as of January 23--26 (\fref{eff}A).
Comparing the models that include only some sources of uncertainty to the ``all'' model, we see that propagating error from the growth rate (which all but one of the studies reviewed did) is absolutely crucial: the middle bar (``GI mean''), which lacks growth-rate uncertainty, is relatively narrow.
In this case, propagating error from the mean generation interval has negligible effect compared to propagating the uncertainty in $r$.
Uncertainty in the generation-interval dispersion also has important effects as it determines the functional form of the relationship between $r$ and \Ro (compare ``growth rate + GI mean'' with ``all'').
For example, reducing the dispersion parameter $\kappa$ from 1 (assuming exponentially distributed generation intervals) to 0 (assuming fixed generation intervals) changes the $r$--\Ro\ relationship from linear to exponential, therefore increasing the sensitivity of \Ro estimates to $r$ and $\bar G$.

@@

We also compare the estimates of \Ro across different studies by 
replacing their values of $r$, $\bar G$, and $\kappa$ with our pooled estimates ($\mu_r$, $\mu_G$, and $\mu_\kappa$, respectively) one at a time and recalculating the basic reproductive number \Ro (\fref{R0}).
This procedure allows us to assess the sensitivity of the estimates of \Ro across appropriate ranges of uncertainties.
We find that incorporating uncertainties one at a time increases the width of the confidence intervals in all but 7 cases.
We estimate narrower confidence intervals for Study 3, Study 6, and Study 7 when we account for proper uncertainties in the generation-interval dispersion because they assume a narrow generation-interval distribution (compare ``base'' with ``GI variation'');
when higher values of $\kappa$ are used, their estimates of \Ro become less sensitive to the values of $r$ and $\bar G$, giving narrower confidence intervals.
We estimate narrower confidence intervals for Study 5 and Study 7 when we account for proper uncertainties in the mean generation interval (compare ``base'' with ``GI mean'') because the range of uncertainty in the mean generation interval $\bar G$ they consider is much wider than the pooled range (\fref{assumption}).
Substituting the reported $r$ or $\bar G$ from Study 1 with our pooled estimates give narrower confidence intervals for similar reasons.

@@

Estimating the basic reproductive number \Ro is crucial for predicting the course of an outbreak and planning intervention strategies.
Here, we use a gamma approximation \cite{park2019practical} to decompose \Ro estimates into three key quantities ($r$, $\bar G$, and $\kappa$) and apply a multilevel Bayesian framework to compare estimates of \Ro for the novel coronavirus outbreak.
Our results demonstrate the importance of accounting for uncertainties associated with the underlying generation-interval distributions, including uncertainties in the amount of dispersion in the generation intervals:
our analysis of individual studies shows that assuming too narrow a generation-interval distribution can make the estimate of \Ro overly sensitive to the estimates of the exponential growth rate $r$.

Of the seven studies that we reviewed, two of them directly fit their models to cumulative number of confirmed cases.
This approach can be appealing because of its simplicity and apparent robustness, but fitting a model to cumulative incidence instead of raw incidence can both bias parameters and give overly narrow confidence intervals, if the resulting non-independent error structures are not taken into account \cite{ma2014estimating, king2015avoidable}.
Naive fits to cumulative incidence data should therefore be avoided.
